---
title: "Assignment1"
author: "RyanOConnor"
date: "9/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, error = T}
library(tidyverse)
##nested for loop that iterates through all combinations of Year_Q for Electric
years <- 2017:2021
quarters <- 1:4
type <- "Electric"

pge_17_21_Electric <- NULL

for(quarter in quarters) {
  for(year in years){
    filename <-
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
    print(filename)
    temp <- read_csv(filename)
    pge_17_21_Electric <- rbind(pge_17_21_Electric,temp)
    # Note rbind requires field names to be consistent for every new thing that you add.
    saveRDS(pge_17_21_Electric, "pge_17_21_Electric.rds")
  }}
#The error in this outpt is expected and tolerated because of the lack of 2021_Q3 and 2021_Q4 CSVs. This error is remedied in the knitr process by annotating the beginning of the for loop chunks with {r, error = T} to instruct the knitr process to ignore the error. 
```

```{r, error = T}

years <- 2017:2021
quarters <- 1:4
type <- "Gas"

pge_17_21_Gas <- NULL
##nested for loop that iterates through all combinations of Year_Qn for Electric

for(quarter in quarters) {
  for(year in years){
    filename <-
      paste0(
        "PGE_",
        year,
        "_Q",
        quarter,
        "_",
        type,
        "UsageByZip.csv"
      )
    print(filename)
    temp <- read_csv(filename)
    pge_17_21_Gas <- rbind(pge_17_21_Gas,temp)
    # Note rbind requires field names to be consistent for every new thing that you add.
    saveRDS(pge_17_21_Gas, "pge_17_21_Gas.rds")
  }}
#The error in this outpt is expected and tolerated because of the lack of 2021_Q3 and 2021_Q4 CSVs. This error is remedied in the knitr process by annotating the beginning of the for loop chunks with {r, error = T} to instruct the knirt process to ignore the error. 
```

```{r}
library(tidyverse)
library(zoo)

pge_elec_filter <- filter(pge_17_21_Electric, CUSTOMERCLASS %in% c("Elec- Residential","Elec- Commercial"))
#Creating an object with only Residential and Commercial Electrical data from the bound object containing all Electric data from 17-12

pge_gas_filter <- filter(pge_17_21_Gas, CUSTOMERCLASS %in% c("Gas- Residential","Gas- Commercial"))
#Same as above for gas data

pge_elec_filter["kBTUs"] <- pge_elec_filter$TOTALKWH * 3412.14/1000
#Conversion from KWH to kBTUs and creating an additional field for kBTUs

pge_gas_filter["kBTUs"] <- pge_gas_filter$TOTALTHM * 99976.1/1000
#Conversion from THERM to kBTUs and creating an additional field for kBTUs

pge_elec_select <-
  select(
    pge_elec_filter,
    -c(COMBINED, AVERAGEKWH, TOTALKWH, AVERAGEKWH)
  )
pge_gas_select <-
  select(
    pge_gas_filter,
    -c(COMBINED, AVERAGETHM, TOTALTHM, AVERAGETHM)
  )
#Removing irrelevant fields so that gas and electric have a similar scheme and can be bound
pge_gas_elec_bound <- 
  rbind(
    pge_gas_select, pge_elec_select)
#Binding gas and electric together, now that the field schema match

pge_gas_elec_bound["Date"] <- "0"
pge_gas_elec_bound$Date <- as.yearmon(paste(pge_gas_elec_bound$YEAR, pge_gas_elec_bound$MONTH), "%Y %m")
#Inserting a Date field that is correctly formatted as a date-type so that the x-axis can represent all months and years together in order. Unsure if ggplot has a way to do this, but I used as.yearmon in the Zoo package, which seemed to work just fine

pge_res_bound <- filter(pge_gas_elec_bound, CUSTOMERCLASS %in% c("Gas- Residential","Elec- Residential"))
pge_comm_bound <- filter(pge_gas_elec_bound, CUSTOMERCLASS %in% c("Elec- Commercial","Gas- Commercial"))
#Filtering so that residential and commercial get their own objects and can be easily visualized independently 
```


```{r}
library(tidyverse)
library(plotly)
library(lubridate)

pge_res_chart <-
  pge_res_bound %>%
  ggplot() +
  geom_bar(
    aes(
      x = Date %>% factor(),
      y = kBTUs,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  theme(axis.text.x = element_text(angle = 75))+
  labs(
    x = "Month",
    y = "kBTUs",
    title = "PG&E Territory Residential Energy Usage, 2017-2021",
    fill = "Electricity Type"
)
    pge_res_chart
#added a theme to the x-axis to angle the text of the axis labels at 75 deg
```

```{r}
library(tidyverse)
library(plotly)
library(lubridate)

pge_comm_chart <-
  pge_comm_bound %>%
  ggplot() +
  geom_bar(
    aes(
      x = Date %>% factor(),
      y = kBTUs,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  theme(axis.text.x = element_text(angle = 75))+
  labs(
    x = "Month",
    y = "kBTUs",
    title = "PG&E Territory Commercial Energy Usage, 2017-2021",
    fill = "Electricity Type"
)
    pge_comm_chart
```

Comment on any observable changes in evergy consumption that may be attributable to the COVID-19 Pandemic:
  
  When looking at the PG&E territorial residential energy use, there is very little appreciable difference between the Pre-Covid (2019) and COVID (2020-21) states in terms of gas energy consumpiton, and only a minor increase in home electricity useage during the Summer of 2020 as compared to the summer of 2019. However, there is a very noticable decline in the energy consumption in the commercial sector in the Spring and Summer of 2020 as compared to 2019, followed by an increase in the Spring and Summer of 2021 above even the "normal" pre-COVID levels of 2019 and prior.The decline in commercial energy consumption in 2020 is intuitively linked to the reduced commercial mobility of society. However, the increase in 2021 is surprising, considering the lingering restrictions in the Bay Area due to COVID, as well as the increase above Pre-COVID levels. One potential explanation for this is a "catch-up," with both consumers and businesses making up for lost time and seeking to engage in more social activity follow the COVID "burnout." What is interesting and notable is that, while intutitvely the proportion of time spent at home increased in 2020, the gross demand of residential energy did not increase significantly, but largely remained the same. This seems to indicate that staying at home more does not necessarily equate to using more energy at home.
  
  Key assumptions that have been made in this analysis of 2019 vs 2020/2021 include the assumption that energy consumption is driven by social mobility of people and the resulting commercial/residential balance of people's time and energy demand. What this analysis does not account for is the potential climactic demand for energy (i.e. if it is warmer in the summer or colder in the winter, there will be greater demand in ebergy) or economic ability to consume energy (i.e. the need to limit energy consumption due to reduced income). Below I've generated 2 plots displaying a zoomed-in view of these energy demand trends, specific to the time period from Jan 2019-Jul 2021
  
```{r}
library(tidyverse)
library(plotly)
library(lubridate)

pge_19_20_21_res_bound <- filter(
pge_res_bound, YEAR %in% c("2019","2020","2021"))

pge_19_20_21_comm_bound <- filter(
pge_comm_bound, YEAR %in% c("2019","2020","2021"))

pge_res19_20_21_chart <-
  pge_19_20_21_res_bound %>%
  ggplot() +
  geom_bar(
    aes(
      x = Date %>% factor(),
      y = kBTUs,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  theme(axis.text.x = element_text(angle = 75))+
  labs(
    x = "Month",
    y = "kBTUs",
    title = "PG&E Territory Residential Energy Usage, 2019-2021",
    fill = "Electricity Type"
)

pge_comm19_20_21_chart <-
  pge_19_20_21_comm_bound %>%
  ggplot() +
  geom_bar(
    aes(
      x = Date %>% factor(),
      y = kBTUs,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  theme(axis.text.x = element_text(angle = 90))+
  labs(
    x = "Month",
    y = "kBTUs",
    title = "PG&E Commercial Residential Energy Usage, 2019-2021",
    fill = "Electricity Type"
)
    
    pge_comm19_20_21_chart
    pge_res19_20_21_chart
```


